PanGraph: scalable bacterial pan-genome graph construction

### 2023-02-23 reviewer_response.tex

\begin{figure}[htb]
      \includegraphics[width=.7\textwidth]{figs_response/hp_vs_kp.pdf}
      \caption{frequency and length distribution (respectively left and right panels) of pancontigs for the \textit{Helicobacter pylori} and \textit{Klebsiella pneumoniae} datasets (top and bottom). For each panel we report the results for pangenome graphs built using minimap2 aligner with standard parameter value and asm20 sensitivity (blue) and with mmseqs2 and null pseudo-energy parameters (orange). Distributions are weighed by pancontig length, so that the area under the curve represents the total pangenome graph size.}
      \label{fig:kp-vs-hp}
\end{figure}

\reviewer{R2.8}{Just to make sure i understand - is it possible for sequence to be ``left out'' in between pancontigs?}
% \Marco{no, every part of the genome must be contained in a pancontig. It can be left out from the consensus if it is considered an insertion.}\\
% \Liam{if they possibly misunderstood, suggests that some statement that all sequence ends up in a pancontig should be added to manuscript. I suspect they are thinking of case of short ($<$100bp) stretches `in between' pancontigs, which (if I understand correctly) will be incorporated as insertions at the end of the consensus alignment of the adjacent pancontig.}
\response{No, every part of the sequence of each genome must be included in some pancontig. I might not necessarily appear in the pancontig consensus sequence if it is present in the alignment as a short ($< \Lthr$) insertion, but it will nonetheless be present in the alignment. In the particular case in which merging two pancontigs would leave overhangs shorter than $\Lthr$, these are included in the alignmen as flanking insertions. This is now explained in SI section I, and mentioned in the main text in section IB (line 103).}

@@

\reviewer{R2.9}{About data accessibility
      \begin{enumerate}
            \item i don't see how to get the list of accessions for section IIB? Saying they were used in the Ding et al paper is not quite enough for me to be sure which ones you used.
            \item for the graph marginalisation section, could the accessions of the strains used be in a supplementary file as well as on github? Github repos are great, but they can be deleted.
      \end{enumerate}}
% \Marco{on the repo we have \href{https://github.com/neherlab/pangraph/blob/master/script/config/accnums.json}{this file} for the list of all accession numbers we use, and \href{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}{this file} for the marginalization pairs. I could create some excel spreadsheet with accession numbers that we could add to the submission?}\\
% \Liam{for sure that's all they want (most biologists don't like json!)}
\responsedraft{We now included in the submission two excel files. ``accession\_numbers.xlsx'' contains five sheets, one per species considered, each with a list of the accession numbers of isolates used in our datasets. ``marginalize\_pairs.xlsx'' contains instead a list of all the pairs used in the graph marginalization section.
      % TODO: archive repository state.
}

\subsubsection*{Minor Issues}

\reviewer{R2.1}{line 13 - it would be good to give a reference for the statement that even closely related genomes can differ widely in gene content.
      I don't know what the first paper to show this was, the earliest i am aware of is from Eduardo Rocha's team, which annpoyingly i can't locate right now, but Figure 2 here shows it \url{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1008866} \cite{touchon2020phylogenetic} - you can be very close in patristic distance and have very different gene repertoire.
      Feel free to replace with a better reference, this si just meant to be helpful.}
% \Marco{I think this is a good reference, fig. 2 is familiar to us. And also fig.2 in \cite{doolittle2009origin} which is a bit older. But not sure which other Rocha paper he is referring to.}\\
% \Liam{I think they mean Figure 5 of \cite{touchon2009organised} (\href{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000344}{here}). Citing Touchon 2009 and 2020 and the Doolittle paper seems fine to me.}
\response{We thank the reviewer for the suggestion. We now cite the suggested article, along with \cite{touchon2009organised,doolittle2009origin}.}

### 2023-02-23 main.tex
update

%\section{Introduction}
During evolution, microbial genomes change by both local mutations and large-scale alterations \cite{arnold2021horizontal}.
Local mutations only change a few nucleotides by substitution, insertion or deletion.
Conversely, large-scale alterations reorganize the sequence, and involve either the homologous recombination of large segments, gene loss or gain, inversions, or mobilisation of genetic elements.
The accumulation of such changes over time complicates comparative genomic analyses of present day isolates.
Homologous recombination is rapid enough that most genes in many bacterial core genomes have distinct phylogenies \cite{sakoparnig2021whole} and even closely related genomes differ dramatically in gene content.
% comment: rewritten to use consistent terminology (I think 'mutation' and 'polymorphism' for most people make them think of local changes only)
%Scalable inference techniques of this complicated relatedness structure would, in part, spur the advance of critical epidemiological surveillance tools, analogous to NextStrain, for bacterial pathogens.

@@

\paragraph*{Merging of homologous sequence.}
If the above alignment step detected homologous stretches between \oldnew{}{the consensus sequences of} two pancontigs, these pancontigs, or parts of them, can be merged.
It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.
We rank each alignment between two pancontigs according to the pseudo-energy
\begin{equation}\label{eq:pseudo-energy}
    E = -\ell + \alpha N_c + \beta N_m
\end{equation}
where $\ell$, $N_c$, and $N_m$ denote the alignment length, number of \emph{pancontigs} created by the merger, and number of polymorphisms per genome in the newly created {pancontig} respectively.
See Fig \ref{fig:visualization}B for a graphical definition for each term.
Additionally, $\alpha$ and $\beta$ are hyperparameters of the algorithm.
Only mergers whose alignment has negative pseudo-energy are performed.
The parameter $\beta$ controls the maximal sequence divergence between the two merger-candidates\oldnew{ and no mergers will be performed if the Hamming distance exceeds $1/\beta$.}{. The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$.}
Importantly, this divergence threshold is applied to the comparison of the consensus sequences of the merger candidates.
\oldnew{}{The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences being merged first.}
The parameter $\alpha$ controls the fragmentation of the graph and imposes a minimal length on the \oldnew{resulting}{} pan-contigs resulting from the merger.
The merger is always performed if \oldnew{the merger}{it} doesn't increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
\oldnew{In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.}{In addition to the fragmentation parameter $\alpha$, there is a further parameter $\Lthr$ that controls the minimal length of pancontigs. Structural variation at a scale smaller than this threshold (i.e. short indels) is stored in pancontig alignments, rather than in the graph structure, and no merges are performed if they have length shorter than this threshold. We set a default value of 100 bp for this parameter.}

>>>>

%\section{Introduction}
During evolution, microbial genomes change by both local mutations and large-scale alterations \cite{arnold2021horizontal}.
Local mutations only change a few nucleotides by substitution, insertion or deletion.
Conversely, large-scale alterations reorganize the sequence, and involve either the homologous recombination of large segments, gene loss or gain, inversions, or mobilisation of genetic elements.
The accumulation of such changes over time complicates comparative genomic analyses of present day isolates.
Homologous recombination is rapid enough that most genes in many bacterial core genomes have distinct phylogenies \cite{sakoparnig2021whole} and even closely related genomes differ dramatically in gene content \oldnew{}{\cite{touchon2020phylogenetic,touchon2009organised,doolittle2009origin}}.
% \cite{touchon2020phylogenetic,doolittle2009origin,touchon2009organised}.
% comment: rewritten to use consistent terminology (I think 'mutation' and 'polymorphism' for most people make them think of local changes only)
%Scalable inference techniques of this complicated relatedness structure would, in part, spur the advance of critical epidemiological surveillance tools, analogous to NextStrain, for bacterial pathogens.

@@

\paragraph*{Merging of homologous sequence.}
If the above alignment step detected homologous stretches between \oldnew{}{the consensus sequences of} two pancontigs, these pancontigs, or parts of them, can be merged.
It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.
We rank each alignment between two pancontigs according to the pseudo-energy
\begin{equation}\label{eq:pseudo-energy}
    E = -\ell + \alpha N_c + \beta N_m
\end{equation}
where $\ell$, $N_c$, and $N_m$ denote the alignment length, number of \emph{pancontigs} created by the merger, and number of polymorphisms per genome in the newly created {pancontig} respectively.
See Fig \ref{fig:visualization}B for a graphical definition for each term.
Additionally, $\alpha$ and $\beta$ are hyperparameters of the algorithm.
Only mergers whose alignment has negative pseudo-energy are performed.
The parameter $\beta$ controls the maximal sequence divergence between the two merger-candidates\oldnew{ and no mergers will be performed if the Hamming distance exceeds $1/\beta$.}{. The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$.}
Importantly, this divergence threshold is applied to the comparison of the consensus sequences of the merger candidates.
\oldnew{}{The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences being merged first.}
The parameter $\alpha$ controls the fragmentation of the graph and imposes a minimal length on the \oldnew{resulting}{} pan-contigs resulting from the merger.
The merger is always performed if \oldnew{the merger}{it} doesn't increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
\oldnew{In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.}{In addition to the fragmentation parameter $\alpha$, there is a further parameter $\Lthr$ that controls the minimal length of pancontigs. Structural variation at a scale smaller than this threshold (i.e. short indels) is stored in pancontig alignments, rather than in the graph structure, and no merges are performed if they have length shorter than this threshold. For instance, when a merge would generate flanking contigs shorter than this threshold, the alignment is extended to include these regions. We set a default value of 100 bp for this parameter.}
